 

### 2021-07-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. 
~~This variation shapes vertical $T_{leaf}$ patterns [Table 1, @fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].~~
As much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al. 2004) was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses through the vertical canopy. 
It has been established that independent of light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy. For example, LMA, photosynthetic capacity [per area] and $gs$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].

@@

Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, by being smaller, hairier, and with higher vein density than shade leaves [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed200;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*]. These traits (Table 1) collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance. In contrast, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks (e.g., lower LMA), but are more likely to get overheated than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. In open canopies where light is comparatively homogeneous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table 1). 
Caratenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin ($VAZ$) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. This is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@TemperaturedependentAdjustmentThermal1996].

Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016a;@niinemetsWithincanopyVariationRate2004;@weerasingheCanopyPositionAffects2014]. Within-canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. Hence, although we would expect elemental traits to influence vertical foliar thermal sensitivity,  @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.

@@

Maximum $gs$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves [Table 2]. However, because of the high $VPD$ and irradiance experienced by sun leaves,  mid-day stomatal depression is prevalent in sun leaves but not in shade leaves.  $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. Higher maximum temperatures influences leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear. Additionally, $T_{opt}$ of $gs$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], but sun leaves show a stronger decrease in $gs$ to rising temperature than shade leaves [@hernandezSimilarTemperatureDependence2020; Carter et al. 2021].

>>>>

Within-canopy, leaf traits vary along the canopy light gradient [*i.e.,* in sun and shade leaves, @casasSunShadeLeaves2011] at a leaf anatomical, structural and biochemical level [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. 
As much of our understanding of trait coordination is based on the leaf economics spectrum concept (Wright et al. 2004) was developed using sun leaves [@chenLeafEconomicsSpectrum2020; @keenanGlobalLeafTrait2016], further research is needed to characterize trait relationships and responses through the vertical canopy. 
It has been established that independent of light environment, traits also vary within species along the ontogenetic trajectory from the understory to the canopy. For example, LMA, photosynthetic capacity (per area) and $g_s$ have been observed to increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000].

@@

Sun leaves in the exposed canopy have traits that help avoid heat and water stress -- for instance, by being smaller, hairier, and with higher vein density than shade leaves [@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006; @zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], and with steeper leaf-angle [@niinemetsAdjustmentFoliageStructure1998, *REF*]. These traits (Table 1) collectively aid heat dissipation and regulate $T_{leaf}$-photosynthetic balance. In contrast, shade leaves have traits that maximize light capture in the diffused-light environment with sunflecks (e.g., lower LMA), but are more likely to get overheated than sun leaves due to greater thermal capacitance [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. In open canopies where light is comparatively homogeneous, exposed leaves are primarily shaped by $T_{air_{max}}$ and $VPD$ stress, rather than solely by light [@mediavillaFoliarPlasticityRelated2019a].

@@

Biochemical protection against foliage light and heat damage also increases with irradiance, and thus tends to be higher in the upper canopy than in the understory (Table 1). 
Caratenoids and xanthophyll cycle pigments -- violaxanthin, antheraxanthin and zeaxanthin ($VAZ$) -- can play a role in antioxidant scavenging and converting excess excitation energy into heat to reduce photoinhibition [@niinemetsPhotosynthesisResourceDistribution2007a; @mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. This is also important for leaf acclimation to higher temperature if high $T_{leaf}$ impairs the photochemical energy dissipation pathway [@havauxTemperaturedependentAdjustmentThermal1996].

Elemental traits, such as $N_{area}$ and $P_{area}$, contribute to activities of high leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes in sun leaves [Table 1, @scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014]. Within-canopy leaf nitrogen distribution is regulated to prevent photoinhibition in shade leaves during sunflecks, and direct light for sun leaves [@kitaoCanopyNitrogenDistribution2018]. Hence, although we would expect elemental traits to influence vertical foliar thermal sensitivity,  @zhuPlasticityPhotosyntheticHeat2018 and @osullivanThermalLimitsLeaf2017 observed no relation between foliar thermal tolerance and total foliar concentrations of N and P.

@@

Maximum $g_s$ increases with light environment, and is thus higher in the sun exposed upper canopy than in subcanopy or understory leaves (Table 2). However, because of the high $VPD$ and irradiance experienced by sun leaves,  mid-day stomatal depression is prevalent in sun leaves but not in shade leaves.  $T_{leaf}$ of sun leaves thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. Higher maximum temperatures and $g_s$ influence leaf thermal sensitivity tolerance [@perezPhotosyntheticHeatTolerances2020], but exact mechanisms remain unclear. Additionally, $T_{opt}$ of $g_s$  did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], but sun leaves show a stronger decrease in $gs$ to rising temperature than shade leaves [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical].
